Index measures for oak decline severity using phenotypic descriptors

### 2020-10-28 manuscript/manuscript.Rmd
manuscript corrections

Phenotyping protocols were developed to specifically measure oak decline severity and were based on a multitude of simple to measure phenotypic descriptors. 
There were a total of 36 phenotypic measurements describing oak decline status which included aspects of tree size, crown condition, the presence of biotic agents and a number of derived composite descriptors.

@@

The suitability of the decline indexes for describing decline severity and type were assessed by comparing decline index scores to manual decline status classifications along with an assessment of descriptor importance and contribution to the decline index models.
Crown condition and trees size descriptors such as composite crown volume contributed positively to the PDI.
Trees with smaller crowns in poor condition had greater PDI values. 
Tree stature and the presence of stem bleeding contributed highly to the DAI, allowing differentiation between trees with AOD and COD syndromes. 
AOD trees had relatively larger stature and the presence of stem bleeding where as COD trees had small stature and stem bleeding absent.

@@

Trees and forests across the globe face threats from a changing climate.
Notably warming and an increase in drought episodes can adversely affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015]. 
Within this context, biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017]. 
Current ‘omics and remote sensing approaches provide powerful tools to document and understand these changes across multiple scales (from cellular to tree to landscape scales), but an important prerequisite is the ability to accurately describe tree health status, so that even subtle changes can be documented [@dungey2018; @hornero2020]. 
Phenotyping, documenting the physical traits of an individual or group, is thus a crucial first step relevant to studies across disciplines and further, it is necessary to understand how these measurements combine to describe health status.

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both long-term predisposing and one-off inciting factors.
Biotic agents can include insect infestation such as the larvae of the bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
Over time, as lesion development continues beneath the bark, external cracks form between the bark plates that are surrounded by blackening stains.
Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of callused tissue on the exterior [@denman2014].
Stem bleeds can also be associated with a range of root pathogens including *Armillaria* attack on the inner stem tissues of oak.
Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].

@@

The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England (Table \@ref(tab:site-information)).
The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
These sites included a variety of woodlands, management histories as well as tree sizes/ages and were made up of mainly *Q. robur*.

@@

The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
In each year of sampling, surveyors underwent refresher training and “calibration” assessments in oak woodland. 
The descriptors collected were based on @innes1990 and @icpforests2016.
Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
The timber height was measured at either where the upper stem diameter was 70mm or where no main stem could be distinguished.
For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval ranges.

@@

Crown radius, canopy closure and bark tap tests for dead stem tissue were measured for each cardinal point.
The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m. 
Canopy closure was recorded as yes or no depending on whether any other crowns were touching the sampling tree.
A high resolution photo was taken of the base to the top of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.

The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
The frequency of *A. biguttatus* exit holes was counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.

@@

All cardinal assessments were averaged to give a single value per tree.
The crown density ($CD$) values collected at Site 1 and 6 were converted to crown transparency ($t = 100 - CD$)
All descriptors collected in cm were converted to mm except for DBH which was converted to m.
The bark tap tests that were positive for hollow stem tissue were converted to present and absent then renamed dead stem tissue.
The canopy closure descriptor was converted from yes and no to present and absent.
Finally, the stem symptoms lengths for active bleeds, black staining and callused wounds were averaged and a frequency per tree given by the total observed.
The application and effect of this adjustment is described and discussed in Sections 3.1 and 4.3 respectively.
After preparation there were a total of 31 raw descriptors for analysis.

@@

A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary Figure S3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary Figure S3b). 
To counter this effect, sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_adjustment_margin$Margin %>% signif(digits = 3)`.
The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary Figure S4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure S4b).

@@

```{r decline-index-plot,fig.cap='Decline indexes for all 174 surveyed trees. The chosen group categories were selected in the field prior to data collection. AOD = acute oak decline; COD = chronic oak decline. Remisson tress were those that were known to have had active bleeds within the previous 3 years.'}
loadd(decline_indexes_plot)

@@

The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure \@ref(fig:descriptor-contributions)a).
Other tree size related descriptors such as bleed prevalence and live crown ratio stem bleeding descriptors contributed secondarily.

@@

AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure \@ref(fig:decline-index-plot)).
The association of crown condition and AOD severity has previously be shown by @brown2016.

The remission trees as a group were shown to be intermediate in decline severity and type to the healthy, AOD and COD trees on both the PDI and DAI index scales.
However, this was also the most varied of the decline status groups.
It is possible that this could reflect the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
Using these decline indexes, remission trees could be more objectively defined in terms of their temporal trajectory within the decline index continuum.

@@

The decline indexes presented here provide a new framework to quantitatively describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
Their basis is on simple phenotypic descriptor measurements and a utilisation of a data driven, unsupervised random forest based machine learning approach.
This has shown the relative contributions that aspects of tree size, crown condition and symptoms such as stem bleeding have on quantitatively defining our concepts of oak decline status.

>>>>

Phenotyping protocols were developed to specifically measure oak decline severity and were based on a comprehensive set of simple to measure phenotypic descriptors. 
A total of 36 phenotypic measurements describing oak decline status included aspects of tree size, crown condition, the presence of biotic agents and a number of derived composite descriptors.

@@

The suitability of the decline indexes for describing decline severity and type were assessed by comparing decline index scores to manual decline status classifications along with an assessment of descriptor importance and contribution to the decline index models.
Crown condition and trees size descriptors such as 'composite crown volume' contributed positively to the PDI.
Trees with smaller crowns in poor condition had greater PDI values. 
Tree stature and the presence of stem bleeding contributed highly to the DAI, allowing differentiation between trees with AOD and COD syndromes. 
AOD trees had relatively larger stature and the presence of stem bleeding where as COD trees had small stature and stem bleeding was absent.

@@

Trees and forests across the globe face threats from a changing climate.
Notably warming and an increase in drought episodes can adversely affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015]. 
Within this context, biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017]. 
Current omics and remote sensing approaches provide powerful tools to document and understand these changes across multiple scales (from cellular to tree to landscape scales), but an important prerequisite is the ability to accurately describe tree health status, so that even subtle changes can be documented [@dungey2018; @hornero2020]. 
Phenotyping, documenting the physical traits of an individual or group, is thus a crucial first step relevant to studies across disciplines and further, it is necessary to understand how these measurements combine to describe health status.

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both long-term predisposing and one-off inciting factors.
Biotic agents can include insect infestation such as larvae of the bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated with Chronic Oak Decline (COD) [@denman2017].
The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
Over time, as lesion development continues beneath the bark, external cracks form between the bark plates that are surrounded by blackening stains.
Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of callused tissue on the exterior [@denman2014].
Stem bleeds can also be associated with a range of root pathogens including *Armillaria* attack on the inner stem tissues of oak.
Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].

@@

The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England (Table \@ref(tab:site-information)).
The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
These sites included a variety of woodlands with diverse management histories, tree sizes/ages and were made up of mainly *Q. robur*.

@@

The phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
In each year of sampling, surveyors underwent refresher training and “calibration” assessments in oak woodland. 
The descriptors collected were based on @innes1990 and @icpforests2016.
Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
Total height, timber height and lowest live crown were measured using a Vertex IV Digital Hypsometer (Haglof, Sweden) to the nearest 0.1m.
The timber height was measured at either where the upper stem diameter was 70mm or where no main stem could be distinguished.
For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval ranges.

@@

Crown radius, canopy closure and bark tap tests for dead stem tissue were measured for each cardinal point.
The crown radius was measured using a Vertex IV Digital Hypsometer attached to the relevant side of the stem at a height of 1.3m. 
Canopy closure was recorded as yes or no depending on whether any other crowns were touching the sampled tree.
A high resolution photo was taken of the base to the top of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.

The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
The frequency of *A. biguttatus* exit holes was counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m from the stem.

@@

All cardinal assessments were averaged to give a single value per tree.
The crown density ($CD$) values collected at Site 1 and 6 were converted to crown transparency ($t = 100 - CD$).
All descriptors collected in cm were converted to mm except for DBH which was converted to m.
The bark tap tests that were positive for hollow stem tissue were converted to present and absent then renamed dead stem tissue.
The canopy closure descriptor was converted from yes and no to present and absent.
Finally, the stem symptoms lengths for active bleeds, black staining and callused wounds were averaged and a frequency per tree given by the total observed.
The application and effect of this adjustment is described and discussed in Sections 3.1 and 4.3 respectively.
After preparation there were a total of 31 raw descriptors for analysis.

@@

A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary Figure S3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary Figure S3b). 
To counter this effect, sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_adjustment_margin$Margin %>% signif(digits = 3)`.
The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able to accurately classify the site locations of the trees (Supplementary Figure S4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure S4b).

@@

```{r decline-index-plot,fig.cap='Decline indexes for all 174 surveyed trees. The decline status categories were manually assigned in the field prior to data collection. AOD = acute oak decline; COD = chronic oak decline. Remisson tress were those that were known to have had active bleeds within the previous 3 years.'}
loadd(decline_indexes_plot)

@@

The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure \@ref(fig:descriptor-contributions)a).
Other tree size related descriptors such as bleed prevalence, live crown ratio and stem bleeding descriptors contributed secondarily.

@@

AOD severity was associated with overall decline severity in trees having both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure \@ref(fig:decline-index-plot)).
The association of crown condition and AOD severity has previously been shown by @brown2016.

The remission trees as a group were shown to be intermediate in decline severity and type to the healthy, AOD and COD trees on both the PDI and DAI index scales.
However, this was also the most varied of the decline status groups.
It is possible that this could reflect the potential for trees to recover from AOD but this is speculative and it is not possible to test this suggestion further with current observations.
Using these decline indexes, remission trees could be more objectively defined in terms of their temporal trajectory within the decline index continuum.

@@

The decline indexes presented here provide a new framework to quantitatively describe the phenotypic oak decline spectrum using intuitive measures of severity (PDI) and type (DAI).
Their basis is on simple phenotypic descriptor measurements and a utilisation of a data driven, unsupervised random forest based machine learning approach.
This has shown the relative contributions that aspects of tree size, crown condition and symptoms such as stem bleeding have on quantitatively defining our concepts of oak decline status.
